Semantic Linearity in a Lazy Optimising Compiler

### 2025-10-07 popl26-accepted-before-rebuttal.tex

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e', \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]

     \infer[]
     {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
     % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
     % \Downarrow \Theta_n : v_n \quad
     (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
     e_i} \mid e' )\Downarrow (\Theta'
     \mid v) : \tau
     }
     {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
     \Downarrow (\Theta' \mid v) : \tau}

   \end{array}
 \]
 \caption{Instrumented Semantics for $\lambda_\Delta^\pi$\label{fig:instrsemapp}}
 \end{figure}

 Following~\cite{Gunter1993APA,cite:linearhaskell}, we also consider
 partial derivations of our semantics in order to be able to state a
 progress result. Given our set of rules defining judgments of the
 form $a \Downarrow b$ with ordered premises, we define a total
 derivation of $a \Downarrow b$ as a tree in the standard way. A
 partial derivation $a \Downarrow?$ is either a root labelled with $a
 \Downarrow?$ or an application of a rule matching $a$ on the left of
 the evaluation arrow, where exactly one of the premises $a'
 \Downarrow?$ has a partial  derivation, all the premises to the left
 of $a' \Downarrow?$ have a total derivation, and the premises to the
 right of $a'\Downarrow?$ are not known yet.
 %

 By construction, in a partial derivation, there is only one premise
 $b\Downarrow?$ with no sub-derivation. We call $b$ the head of such a
 partial derivation.
We write $a \Downarrow^* b$ for the relation which holds when $b$ is
the head of some partial derivation with root $a \Downarrow?$.  We
write $a \Downarrow^* b$  the partial evaluation relation.

Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation
environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
$\Theta\downarrow^1_e$ refers to the expressions bound to linear variables in
the runtime environment and $\Theta\downarrow^1_\sigma$ refers to the types of
those expressions.

 \begin{definition}[Environment Expansion]
   We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
   \[
     \begin{array}{lcl}
       \left[ \cdot \mid e \right] & \triangleq & e\\
       \left[ \Theta , \xl = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda \xl . e)~e']\\
       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
    \end{array}
   \]
 \end{definition}

We note that for the treatment of recursive let groups, environment
expansion needs to happen for the entire group of let-bound
variables. This is easily achieved by extending the runtime
environment with information about recursive let groupings, which we
omit for the sake of presentation.

 \begin{definition}[Well-typed State]
   We write $\G;\D \vdash \Theta \mid e : \tau,\Sigma$ to denote the following typing judgment:

   \[
\G ; \D \vdash [ \Theta \mid (e,\mathit{terms}(\Sigma))] : \tau \otimes \bigotimes(\Sigma)
   \]
   where $\mathit{terms}(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ stands for the tuple $(e_1 ,( \dots , (e_n , ())))$ and
   $\bigotimes(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ for $\sigma_1 \otimes ( \dots (\sigma_n \otimes ()))$.
 \end{definition}

 \begin{lemma}~\label{lem:unwrapenv}
   $\G ; \D \vdash [ \Theta \mid e] : \sigma$
     iff $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
   \end{lemma}
\begin{proof}
By structural induction on the environment expansion definition in the
forward direction and by a straightforward induction on $\Theta$ in the
converse direction. We show only the forward direction.
\begin{description}
%\item[Case:] $\left[\cdot \mid e\right] \triangleq e$ Trivial
\item[Case:] $\left[\Theta, \xl = y \mid e\right] \triangleq \left[\Theta \mid (\lambda \xl. e)~y\right]$
\begin{tabbing}
    (1) $\G; \D \vdash \left[ \Theta, \xl = y \mid e \right] : \phi$\\
    (2) $\G; \D \vdash \left[ \Theta \mid (\lambda \xl. e)~y \right] : \phi$\` by def.\\
    (3) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash ((\lambda \xl. e)~y, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
    (4) $Let~\D_1,\D_2,\D_3 = \D, \Theta\uparrow^1 $\\
    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
    (6) $\G, \Theta\uparrow^\omega; \D_2 \vdash y : \s$\` by inv\\
    (7) $\G, \Theta\uparrow^\omega; \D_1 \vdash (\lambda \xl. e) : \s \to \phi $\` by inv\\
    (8) $\G, \Theta\uparrow^\omega; \D_1, \xl \vdash e : \phi $\` by inv\\
    (9) $\G, \Theta\uparrow^\omega; \D_1, \D_2, \xl \vdash (e, y) : \phi \otimes \sigma$\` by pair\\
    (10) $\G, \Theta\uparrow^\omega; \D, \xl \vdash (e, y, \Theta\downarrow^1_e) : \phi \otimes \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
\end{tabbing}
\item[Case:] $\left[\Theta, \xD = e' \mid e\right] \triangleq \left[\Theta \mid \llet{\xD = e'}{e}\right]$
\begin{tabbing}
    (1) $\G; \D' \vdash \left[ \Theta, \xD = y \mid e \right] : \phi$\\
    (2) $\G; \D' \vdash \left[ \Theta \mid \llet{\xD = e'}{e} \right] : \phi$\` by def.\\
    (3) $\G, \Theta\uparrow^\omega; \D', \Theta\uparrow^1 \vdash (\llet{\xD = e'}{e}, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
    (4) $Let~\D,\D_2,\D_3 = \D', \Theta\uparrow^1 $\\
    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
    (6) $\G, \Theta\uparrow^\omega; \D \vdash e' : \s$\` by inv on let\\
    (7) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2 \vdash e : \phi $\` by inv on let\\
    (8) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2,\D_3 \vdash (e, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
\end{tabbing}
\end{description}
\end{proof}

\subsection{Bisimulation}

We relate our two operational semantics through a bisimulation,
allowing us to transfer type safety of the instrumented semantics to
the lazy big-step semantics. The relation essentially allows the
execution environment to treat some unrestricted bindings
of the natural semantics as linear.

\begin{definition}[Bisimulation]
We define the relation $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e' : \tau,\Sigma)$ between a well-typed state and an ordinary state of
the lazy big-step semantics if $e = e'$ and $\Theta'{\uparrow^\omega}
\leq \Theta$.
\end{definition}

\begin{lemma}~\label{lem:naturaltoinstr}
  \begin{itemize}
    \item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e : \tau,\Sigma)$ such that $\Theta : e \Downarrow \Theta'' : v$
there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
\tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
(\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
\vdash \Theta''' \mid v : \tau,\Sigma)$
\item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e : \tau,\Sigma)$ such that $\Theta : e \Downarrow^* \Theta'' : v$
there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
\tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
(\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
\vdash \Theta''' \mid v : \tau,\Sigma)$
\end{itemize}
\end{lemma}
\begin{proof}
By induction on the given operational semantics rules.
  \end{proof}

\begin{lemma}~\label{lem:instrtonatural}
  \begin{itemize}
\item  For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
(\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
such that $\Theta : e \Downarrow \Theta''' : v$ and
$\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
\tau,\Sigma)$
\item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
(\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
such that $\Theta : e \Downarrow^* \Theta''' : v$ and
$\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
\tau,\Sigma)$
   \end{itemize}
 \end{lemma}
\begin{proof}
By induction on the given operational semantics rules.
  \end{proof}

\section{Type Safety Proofs}\label{app:proofs}

In the following we write $\Mapsto$ to stand for
$\Rightarrow_\mathsf{WHNF}$ and $\Rrightarrow$ for $\Rightarrow_\mathsf{NWHNF}$.

\subsection{Auxiliary Lemmas}
The following lemmas presuppose the variable names across the typing
contexts are distinct.

  \begin{lemma}[Linear to $\Delta$-bound]~\label{lem:onedelta-app}

    \begin{enumerate}

      \item If $\G; \D',\xl \vdash e :\vp$
        then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ (with $\Delta$ fresh).
      \item If $\G ; \D' , \xl \vdash_{alt} \rho \rightarrow e
        :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
        $\G[\D/x],\xD;\D,\D' \vdash_{alt} \rho \rightarrow e
        :^z_{\Delta_s} \sigma \Rightarrow \rho$ (with $\Delta$ fresh).

      \end{enumerate}
\end{lemma}
\begin{proof}
Straightforward induction on the structure of the given derivation,
reconstructing the original derivation and applying rule
$\mathit{Var}_\Delta$ where rule $\mathit{Var}_1$ was applied on
$x$ in the original derivation.
\end{proof}
\onedelta*

\begin{proof}
By Lemma~\ref{lem:onedelta-app}(1).
\end{proof}

\begin{lemma}[$\Delta$-bound to Linear]~\label{lem:deltaone-app}

  \begin{enumerate}
 \item If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
   then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
 \item If $\G ,\x[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} \rho \rightarrow e
   :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
   $\G[x/\irr{\D}];\D',\xl \vdash_{alt} \rho \rightarrow e
   :^z_{\Delta_s} \sigma \Rightarrow \rho$.
  \end{enumerate}
\end{lemma}
\begin{proof}
Straightforward induction on the structure of the given derivation,
reconstructing the original derivation and applying rule
$\mathit{Var}_1$ where rule $\mathit{Var}_\D$ was applied on
$x$ in the original derivation.
\end{proof}

\deltaone*
\begin{proof}
By Lemma~\ref{lem:deltaone-app}(1).
\end{proof}

\begin{lemma}[Unrestricted and $\Delta$-bound]~\label{lem:undelta-app}
  \begin{enumerate}
    \item $\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e :
      \vp$
    \item $\G,\xo;\D \vdash_{alt} \rho \rightarrow e
      :^z_{\Delta_s} \sigma \Rightarrow \rho$ iff
      $\G,\x[\cdot];\D \vdash_{alt} \rho \rightarrow e
   :^z_{\Delta_s} \sigma \Rightarrow \rho$
  \end{enumerate}
\end{lemma}
\begin{proof}
Straightforward induction on the structure of the given derivation,
noting that rules $\mathit{Var}_\Delta$ and $\mathit{Var}_\omega$
coincide when $\Delta$ is empty.
\end{proof}

\undelta*

\begin{proof}
By Lemma~\ref{lem:undelta-app}.
\end{proof}

\subsection{Irrelevance\label{sec:proof:irrelevance}}

\input{language-v4/proofs/WHNFConvSoundness}

\subsection{Substitution Lemmas\label{sec:proof:substitution-lemmas}}

\input{language-v4/proofs/LinearSubstitutionLemma}

\input{language-v4/proofs/UnrestrictedSubstitutionLemma}

\input{language-v4/proofs/DeltaSubstitutionLemma}

\subsection{Type Preservation\label{sec:proof:type-preservation}}

% \input{language-v4/proofs/TypePreservationTheorem}
\begin{theorem}[Type Preservation]\label{thm:typepres}
  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
  \tau , \Sigma$ or $\G ; \D \vdash (\Theta \mid e) \Downarrow^* (\Theta' \mid v) :
  \tau , \Sigma$.
  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$

\end{theorem}

\begin{proof}
By structural induction on the instrumented operational semantics. We
omit many trivial cases.
\begin{description}
    \item[Case:] $\Gamma; \Delta \vdash (\Theta, \xl = y \mid x) \Downarrow (\Theta' \mid v) : \s, \Sigma$
\begin{tabbing}
    (1) $\G; \D \vdash \Theta, \xl = y \mid x : \s, \Sigma$\\
    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1, \xl \vdash (x, y, \Theta{\downarrow}^1_e, \Sigma) : \s \otimes \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\G';\D_1,\D_2,\D_3 = \G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1$\` s.t. $(\D_2,\D_3)$ types $(\Theta{\downarrow}^1_e,\Sigma)$\\
    (4) $\G'; \xl \vdash x : \s$\\
    (5) $\G'; \D_1 \vdash y : \s$\\
    (6) $\G'; \D_1, \D_2, \D_3 \vdash (y, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\\
    (7) $\G; \D \vdash \Theta \mid y : \s, \Sigma$\`by well-typed state\\
    (8) $\G; \D \vdash \Theta' \mid v : \s, \Sigma$\` by induction hypothesis\\
\end{tabbing}
    \item[Case:] $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) \Downarrow (\Theta', x{:}_{\D_a}\s = v \mid v) : \s, \Sigma$
\begin{tabbing}
    (1) $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) : \s, \Sigma$\\
    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash \llet{x{:}_{\D'} = e}{(x, \Theta{\downarrow}^1_e, \Sigma)} : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
